Appendix 3: Estimating species interactions

### 2015-08-04 Harris_2015/manuscript-materials/draft.tex
almost done

\documentclass[12pt,]{article}
\usepackage[compact]{titlesec}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\begin{spacing}{1.9}
\begin{flushleft}
\emph{Running head:} Species interactions in Markov networks

\textbf{Title: Estimating species interactions from observational data
with Markov networks}

@@

\setlength{\parindent}{1cm} \setlength{\parskip}{6pt}

@@

If nontrophic species interactions, such as competition, are important
drivers of community assembly, then ecologists might expect to see their
influence in our data sets (MacArthur 1958, Diamond 1975). Despite
decades of work and several major controversies, however (Lewin 1983,
Strong et al. 1984, Gotelli and Entsminger 2003, Connor et al. 2013),
existing methods for detecting competition's effects on community
structure are unreliable (Gotelli and Ulrich 2009). In particular,
species' effects on one another can become lost in the complex web of
direct and indirect interactions in real assemblages. For example, the
competitive interaction between the two shrub species in Figure 1A can
become obscured by their shared tendency to occur in unshaded areas
(Figure 1B). If this sort of effect is common, then the vast majority of
ecologists' methods, which rely on test statistics describing the total
number of co-occurrences or overall correlation between putatively
interacting species (Diamond 1975, Strong et al. 1984, Gotelli and
Ulrich 2009, Veech 2013, Pollock et al. 2014) will not generally provide
much evidence regarding species' direct effects on one another.

While competition doesn't reliably leave the expected pattern in
community structure (a low co-occurrence rate at the landscape level),
it nevertheless does leave a signal in the data (Figure 1C).
Specifically, \emph{among shaded sites}, there will be a deficit of
co-occurrences between the competing shrub species, and \emph{among
unshaded sites}, there will also be such a deficit. More generally, we
can obtain much better estimates of the association between two species
from their conditional relationships (i.e.~by controlling for other
species in the network) than we could get from their overall
co-occurrence rates. This kind of precision is difficult to obtain from
null models, which begin with the assumptions that all the pairwise
interactions are zero and thus don't need to be controlled for.
Nevertheless, null models have dominated this field for more than three
decades (Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009).

In this paper, following recent work by Azaele et al. (2010) and Fort
(2013), I show that Markov networks (undirected graphical models also
known as Markov random fields; Murphy 2012) can provide a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observed presence-absence
matrices. Markov networks have been used in many scientific fields for
decades in similar contexts, from physics (where nearby particles
interact magnetically with one another; Cipra 1987) to spatial
statistics (where adjacent grid cells interact with one another; Harris
1974, Gelfand et al. 2005). While community ecologists explored some
related approaches in the 1980's (Whittam and Siegel-Causey 1981),
computational limitations had previously imposed severe approximations
that produced unintelligible results (e.g. ``probabilities'' greater
than one; Gilpin and Diamond 1982). Now that it is computationally
feasible to fit these models exactly, this kind of approach is worth a
second look.

The rest of the paper proceeds as follows. First, I discuss how Markov
networks operate and how they can be used to simulate landscape-level
data or to predict the direct and indirect consequences of possible
interaction matrices. Then, using simulated data sets where the ``true''
ecological structure is known, I compare this approach with several
existing methods for detecting species interactions. Finally, I discuss
opportunities for extending the approach presented here to larger
problems in community ecology.

@@

\paragraph{Conditional relationships and Markov
networks.}\label{conditional-relationships-and-markov-networks.}

Ecologists are often interested in learning about direct interactions
between species, controlling for the indirect influence of other
species. In statistical terms, this implies that ecologists want to
estimate \emph{conditional} (``all-else-equal'') relationships, rather
than \emph{marginal} (``overall'') relationships. The example with the
shrubs and trees in Figure 1 shows how the two measures can have
opposite signs, and suggests that conditional relationships are more
relevant for drawing inferences about species interactions
(e.g.~competition).

Markov networks define species interactions in conditional terms, and
use these conditional relationships to determine whether a given
combination of co-occurring species is consistent with their
interactions. Given a set of binary outcomes for each species (i.e.~0
for absence and 1 for presence), the Markov network defines the relative
probability of observing a given presence-absence vector, \(\vec{y}\),
as

\centering
\(p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).\)
\raggedright

Here, \(\alpha_{i}\) is an intercept term determining the amount that
the presence of species \(i\) contributes to the log-probability of
\(\vec{y}\); it directly controls the prevalence of species \(i\).
Similarly, \(\beta_{ij}\) is the amount that the co-occurrence of
species \(i\) and species \(j\) contributes to the log-probability; it
controls the probability that the two species will be found together
(Figure 2A, Figure 2B). Because the relative probability of a
presence-absence vector increases when positively-associated species
co-occur and decreases when negatively-associated species co-occur, the
model tends to produce assemblages that have many pairs of
positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect).

A major benefit of Markov networks is the fact that the conditional
relationships between species can be read directly off the matrix of
\(\beta\) coefficients (Murphy 2012). For example, if the coefficient
linking two mutualist species is \(+2\), then---all else equal---the
odds of observing either species increase by a factor of \(e^2\) when
its partner is present (Murphy 2012). In this way, the \(\beta\)
coefficients behave very similarly to the coefficients in a logistic
regression model (Lee and Hastie 2012).

Of course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from
this expectation. For this reason, it is important to consider how
coefficients' effects propagate through the network, as discussed below.

Estimating the marginal relationships predicted by a Markov network is
more difficult than estimating conditional relationships, because doing
so requires absolute probability estimates. Turning the relative
probability given by Equation 1 into an absolute probability entails
scaling by a normalizing constant, \(Z\), so that that the probabilities
of all possible assemblages will sum to one (bottom of Figure 2B).
Calculating \(Z\) exactly, as is done in these analyses, quickly becomes
infeasible as the number of species increases: with \(2^N\) possible
assemblages of \(N\) species, the number of bookkeeping operations
required spirals exponentially into the billions and beyond. Numerous
techniques are available for keeping the calculations tractable,
e.g.~via analytic approximations (Lee and Hastie 2012) or Monte Carlo
sampling (Salakhutdinov 2008), but they are beyond the scope of this
paper. Alternatively, as noted below, some common linear and generalized
linear methods can also be used as computationally efficient
approximations to the full network.

@@

In the previous two sections, the values of \(\alpha\) and \(\beta\)
were known. In practice, however, ecologists will often need to estimate
these parameters from co-occurrence data (i.e.~from a matrix of ones and
zeros indicating which species are present at which sites). When the
number of species is reasonably small, one can compute exact maximum
likelihood estimates for all of the \(\alpha\) and \(\beta\)
coefficients given a presence-absence matrix by optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces (Murphy 2012),
ensuring that this procedure will always converge on the global maximum.
This maximum represents the unique combination of \(\alpha\) and
\(\beta\) coefficients that would be expected to produce exactly the
observed co-occurrence frequencies (i.e.~it matches the sufficient
statistics of the model distribution to the sufficient statistics of the
data). For the analyses in this paper, I used the \emph{rosalia} package
(Harris 2015a) for the R programming language (R Core Team 2015) to
define the objective function and gradient in R code and then find the
optimum.

@@

In order to compare different methods, I simulated two sets of
landscapes using known parameters.

The first set of simulated landscapes included the three competing
species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling exactly from a probability
distribution defined by the interaction coefficients in that figure
(Appendix A). Each of the methods described below was then evaluated on
its ability to correctly infer that the two shrub species competed with
one another, despite their frequent co-occurrence.

I also simulated a second set of landscapes using a stochastic community
model based on generalized Lotka-Volterra dynamics, as described in the
Appendix. In these simulations, each species pair was randomly assigned
to either compete for a portion of the available carrying capacity
(negative interaction) or to act as mutualists (positive interaction).
In these simulations, mutualism operated by mitigating the effects of
intraspecific competition on each partner's death rate. For these
analyses, I simulated landscapes with up to 20 species and 25, 200, or
1600 sites (50 replicates per landscape size; see Appendix).

@@

I used Markov networks and several other techniques for determining the
sign and strength of the associations between pairs of species (Appendix
B).

For the Markov networks, I used the rosalia package (Harris 2015a), as
described above. Given the large number of parameters associated with
some of the networks to be estimated, I regularized the likelihood using
a logistic prior distribution with a scale of 2 on the \(\alpha\) and
\(\beta\) terms. This convex prior distribution has a similar shape to
the Laplace distribution used in LASSO regularization (especially in the
tails), but does not force the estimates to be exactly zero.

I also evaluated five alternative methods from the existing literature
to calibrate the performance of the Markov network as a method for
estimating species interactions. The first two alternative interaction
measures were the sample correlations and the partial correlations
between each pair of species' data vectors on the landscape (Albrecht
and Gotelli 2001, Faisal et al. 2010). Because partial correlations are
undefined for landscapes with perfectly-correlated species pairs, I used
a regularized estimate based on James-Stein shrinkage, as implemented in
the corpcor package's \texttt{pcor.shrink} function with the default
settings ({\textbf{???}}). In the context of non-Gaussian data, the
partial correlation can be thought of as a computationally efficient
approximation to the full Markov network model (Loh and Wainwright
2013).

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network (Lee and Hastie 2012). Following Faisal et al. (2010), I fit
regularized logistic regression models for each species, using the other
species on the landscape as predictors. {[}{[}Observational
presence-absence data does not have enough degrees of freedom to
estimate this many parameters, so I symmetrized the pairwise
correlations by averaging the coefficient for predicting species \(i\)'s
status from species \(j\) and vice versa.{]}{]}

The fourth alternative method, described in Gotelli and Ulrich (2009),
involved simulating possible landscapes from a null model that retains
the row and column sums of the original matrix (Strong et al. 1984).
Using the default options in the Pairs software described in Gotelli and
Ulrich (2009), I simulated the null distribution of scaled C-scores (a
test statistic describing the number of \emph{non}-co-occurrences
between two species). The software then calculated a \(Z\) statistic for
each species pair using this null distribution. After multiplying this
statistic by \(-1\) so that positive values corresponded to facilitation
and negative values corresponded to competition, I used it as another
estimate of species interactions.

The last two methods used the latent correlation matrix estimated by the
BayesComm package ({\textbf{???}}) to evaluate the claim from recent
papers that the correlation coefficients estimated by ``joint species
distribution models'' provide an accurate assessment of species'
pairwise interactions (Pollock et al. 2014, see also Harris 2015b). I
reported both the average correlation between each pair of species'
latent variables across 1000 Monte Carlo samples, as well as the average
\emph{partial} correlation.

@@

For the simulated landscapes based on Figure 1, method evaluation was
fairly qualitative: any method whose test statistic for the two shrubs
indicated a negative relationship passed; other methods failed. For
methods that provide confidence intervals or other inferential
statistics (Appendix), I also reported how often they rejected the null
hypothesis of no interaction.

Because the different methods mostly describe species interactions on
different scales (e.g.~correlations versus \(Z\) scores versus
regression coefficients), I used linear regression through the origin to
rescale the different estimates produced by each method so that they had
a consistent interpretation. After rescaling each method's estimates, I
calculated squared errors between the scaled interaction estimates and
``true'' interaction values for each of the {[}{[}N{]}{]} species pairs
across all the simulated data sets. These squared errors determined the
proportion of variance explained for different combinations of method
and landscape size (compared with a baseline model that assumed all
interaction strengths to be zero).

@@

As shown in Figure 1, the marginal relationship between the two shrub
species was positive---despite their competition for space at a
mechanistic level---due to indirect effects of the dominant tree
species. As a result, the covariance method falsely reported positive
associations for 94\% of the simulated landscapes, and the
randomization-based null model falsely reported such associations 100\%
of the time. The methods for evaluating conditional relationships
(Markov networks, GLMs, and both types of partial covariance estimates),
however, successfully controlled for the indirect pathway via the tree
species and each correctly identified the direct negative interaction
between the shrubs 94\% of the time.

@@

The accuracy of the four evaluated methods varied substantially,
depending on the parameters that produced the simulated communities
(Figure 3). In general, however, there was a consistent ordering\ldots{}

The three methods that estimated marginal relationships among species
and did not explicitly control for indirect species interactions (the
null model, the joint species distribution model's correlation
coefficients, and the sample correlation coefficients) differed from one
another in their \(R^2\) values, but their estimates were surprisingly
similar to one another. In particular, the correlation coefficients were
extremely similar to the estimates from the other two methods
(Spearman's rho = X for the joint species distribution model's
correlation matrix and Y for the null model).

@@

It's important to note that none of these methods can identify the exact
nature of the pairwise interactions (e.g.~which species in a
positively-associated pair is facilitating the other), particularly when
real pairs of species can reciprocally influence one another in multiple
ways simultaneously (Bruno et al. 2003); with compositional data, there
is only enough information to provide a single number describing each
species pair.\\On the other hand, if ecologists are familiar enough with
the natural history of their study systems that they can augment the
likelihood function with a prior distribution or a suitable penalty,
then this information could reduce the effective degrees of freedom to
estimate and real-world results might be even better than those shown in
Figure 3.

Despite these limitations, Markov networks have enormous potential to
improve ecological understanding. For example, Markov networks provide a
simple answer to the question of how competition should affect a
species' overall prevalence, which was a major flashpoint for the null
model debates in the 1980's (Roughgarden 1983, Strong et al. 1984).
Equation 1 can be used to calculate the expected prevalence of a species
in the absence of biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee
and Hastie 2012). Competition's effect on prevalence in a Markov network
can then be calculated by subtracting this value and the observed
prevalence (cf Figure 2D).

Markov networks---particularly the Ising model for binary
networks---have been studied in statistical physics for nearly a century
(Cipra 1987), and the models' properties, capabilities, and limits are
well-understood in a huge range of applications. Modeling species
interactions using the same framework would thus allow ecologists to tap
into an enormous set of existing discoveries and techniques for dealing
with indirect effects, stability, and alternative stable states.
Numerous other extensions are possible: Markov networks can be fit with
a mix of discrete and continuous variables, for example (Lee and Hastie
2012). There are even methods (Whittam and Siegel-Causey 1981,
Tjelmeland and Besag 1998) that would allow the coefficient linking two
species in an interaction matrix to vary as a function of the abiotic
environment or of third-party species that tip the balance between
facilitation and exploitation (Bruno et al. 2003).

>>>>

\documentclass[12pt,]{article}
\usepackage[compact]{titlesec}
\usepackage{fancyhdr}
\usepackage{hanging}
\pagestyle{fancy}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\usepackage{ifthen}
\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\begin{spacing}{1.9}
\begin{flushleft}
\chead{\ifthenelse{\value{page}=1}{}{\textit{Species interactions in Markov networks}}}
\renewcommand{\headrulewidth}{0pt}

\setlength{\parskip}{0pt}

\textbf{Title:} Estimating species interactions from observational data
with Markov networks

@@

@@

\setlength{\parindent}{1em} \setlength{\parskip}{3pt}

If nontrophic species interactions, such as competition, are important
drivers of community assembly, then ecologists might expect to see their
influence in species composition data (MacArthur 1958, Diamond 1975).
Despite decades of work and several major controversies, however (Lewin
1983, Strong et al. 1984, Gotelli and Entsminger 2003, Connor et al.
2013), existing methods for detecting competition's effects on community
structure are unreliable (Gotelli and Ulrich 2009). In particular,
species' effects on one another can become lost in the complex web of
direct and indirect interactions in real assemblages. For example, the
competitive interaction between the two shrub species in Figure 1A can
become obscured by their shared tendency to occur in unshaded areas
(Figure 1B). While ecologists have long known that indirect effects can
overwhelm direct ones at the landscape level (Dodson 1970, Levine 1976),
the vast majority of our methods for drawing inferenes from
observational data do not control for these effects (e.g. Diamond 1975,
Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013, Pollock et al.
2014). To the extent that indirect interactions like those in Figure 1
are generally important (Dodson 1970), existing methods will thus not
generally provide much evidence regarding species' direct effects on one
another. The goal of this paper is to resolve this long-standing
problem.

While competition doesn't reliably leave the expected pattern in
community structure (low co-occurrence rates at the landscape level), it
nevertheless does leave a signal in the data (Figure 1C). Specifically,
\emph{among shaded sites}, there will be a deficit of co-occurrences
between the competing shrub species, and \emph{among unshaded sites},
there will also be such a deficit. More generally, we can obtain much
better estimates of the association between two species from their
conditional relationships (i.e.~by controlling for other species in the
network) than we could get from their overall co-occurrence rates. This
kind of precision is difficult to obtain from null models, which begin
with the assumptions that all the pairwise interactions are zero and
thus don't need to be controlled for. Nevertheless, null models have
dominated this field for more than three decades (Strong et al. 1984,
Gotelli and Ulrich 2009).

Following recent work by Azaele et al. (2010) and Fort (2013), this
paper shows that Markov networks (undirected graphical models also known
as Markov random fields; Murphy 2012) can provide a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observed presence-absence
matrices. Markov networks have been used in many scientific fields for
decades in similar contexts, from physics (where nearby particles
interact magnetically; Cipra 1987) to spatial statistics (where adjacent
grid cells have correlated values; Harris 1974, Gelfand et al. 2005).
While community ecologists explored some related approaches in the
1980's (Whittam and Siegel-Causey 1981), they used severe approximations
that led to unintelligible results (e.g. ``probabilities'' greater than
one; Gilpin and Diamond 1982).

Below, I introduce Markov networks and show how they can be used to
simulate landscape-level data or to make exact predictions about the
direct and indirect consequences of possible interaction matrices. Then,
using simulated data sets where the ``true'' interactions are known, I
compare this approach with several existing methods. Finally, I discuss
opportunities for extending the approach presented here to other
problems in community ecology, e.g.~quantifying the overall effect of
species interactions on occurrence rates (Roughgarden 1983) and
disentangling the effects of biotic versus abiotic interactions on
species composition (Kissling et al. 2012, Pollock et al. 2014).

@@

\paragraph{Markov networks.}\label{markov-networks.}

Markov networks provide a framework for translating back and forth
between the conditional relationships among species (Figure 1C) and the
kinds of species assemblages that these relationships produce. Here, I
show how a set of conditional relationships can be used to determine the
how groups of species can co-occur. Methods for estimating conditional
relationships from data are discussed in the next section.

A Markov network defines the relative probability of observing a given
vector of species-level presences (1s) and absences (0s), \(\vec{y}\),
as

\centering

\(p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).\)

\raggedright
\setlength{\parindent}{1em}

Here, \(\alpha_{i}\) is an intercept term determining the amount that
the presence of species \(i\) contributes to the log-probability of
\(\vec{y}\); it directly controls the prevalence of species \(i\).
Similarly, \(\beta_{ij}\) is the amount that the co-occurrence of
species \(i\) and species \(j\) contributes to the log-probability; it
controls the conditional relationship between two species, i.e.~the
probability that they will be found together, after controlling for the
other species in the network (Figure 2A, Figure 2B). For example,
\(\beta_{ij}\) might have a value of \(+2\) for two mutualists,
indicating that the odds of observing one species are \(e^2\) times
higher in sites where its partner is present than in comparable sites
where its partner is absent. Because the relative probability of a
presence-absence vector increases when positively-associated species
co-occur and decreases when negatively-associated species co-occur, the
model tends---all else equal---to produce assemblages that have many
pairs of positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect).

Of course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from
this expectation. Determining the marginal relationships between species
from their conditional interactions entails summing over the different
possible assemblages (Figure 2B). This becomes intractable when the
number of possible assemblages is large, though several methods beyond
the scope of this paper can be employed to keep the calculations
feasible (Salakhutdinov 2008, Lee and Hastie 2012). Alternatively, as
noted below, some common linear and generalized linear methods can also
be used as computationally efficient approximations to the full network
(Lee and Hastie 2012, Loh and Wainwright 2013).

@@

In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a matrix of ones and zeros indicating which species are present at which
sites). When the number of species is reasonably small, one can compute
exact maximum likelihood estimates for all of the \(\alpha\) and
\(\beta\) coefficients given a presence-absence matrix by optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces (Murphy 2012),
ensuring that this procedure will always converge on the global maximum.
This maximum represents the unique combination of \(\alpha\) and
\(\beta\) coefficients that would be expected to produce exactly the
observed co-occurrence frequencies on average (i.e.~it matches the
sufficient statistics of the model distribution to the sufficient
statistics of the data; Murphy 2012). I used the rosalia package (Harris
2015a) for the R programming language (R Core Team 2015) to optimize the
Markov network parameters. The package was named after Santa Rosalia,
the patron saint of biodiversity, whose supposedly miraculous healing
powers played an important rhetorical role in the null model debates of
the 1970's and 1980's (Lewin 1983).

@@

In order to compare different methods, I simulated two sets of
landscapes using known parameters. The first set included the three
competing species shown in Figure 1. For each of 1000 replicates, I
generated a landscape with 100 sites by sampling from a probability
distribution defined by the figure's interaction coefficients (Appendix
1). Each of the methods described below was then evaluated on its
ability to correctly infer that the two shrub species competed with one
another, despite their frequent co-occurrence.

I also simulated a second set of landscapes using a stochastic community
model based on generalized Lotka-Volterra dynamics, as described in
Appendix 2. In these simulations, each species pair was randomly
assigned to either compete for a portion of the available carrying
capacity (negative interaction) or to act as mutualists (positive
interaction). Here, mutualisms operate by mitigating the effects of
intraspecific competition on each partner's death rate. For these
analyses, I simulated landscapes with up to 20 species and 25, 200, or
1600 sites (50 replicates per landscape size; see Appendix 2).

@@

I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).

First, I used the rosalia package (Harris 2015a) to fit Markov newtork
models, as described above. For the analyses with 20 species, I added a
very weak logistic prior distribution on the \(\alpha\) and \(\beta\)
terms with scale 2 to ensure that the model estimates were always
finite. The bias introduced by this prior should be small: the 95\%
credible interval on \(\beta\) only requires that one species' effect on
the odds of observing a different species to be less than a factor of
1500 (which is not much of a constraint). The logistic distribution was
chosen because it is convex and has a similar shape to the Laplace
distribution used in LASSO regularization (especially in the tails), but
unlike the Laplace distribution it is differentiable everywhere and does
not force any estimates to be exactly zero. To confirm that this
procedure produced stable estimates, I compared its estimates on 50
bootstrap replicates (Appendix 4).

I also evaluated five alternative methods from the existing literature
(plus a novel variant of the fifth method). The first two alternative
interaction measures were the sample correlations and the partial
correlations between each pair of species' data vectors on the landscape
(Albrecht and Gotelli 2001, Faisal et al. 2010). In the context of
non-Gaussian data, the partial correlation can be thought of as a
computationally efficient approximation to the full Markov network model
(Loh and Wainwright 2013). This sort of model is very common for
estimating relationships among genes and gene products (Friedman et al.
2008). Because partial correlations are undefined for landscapes with
perfectly-correlated species pairs, I used a regularized estimate based
on James-Stein shrinkage, as implemented in the corpcor package's
\texttt{pcor.shrink} function with the default settings (Schäfer et al.
2014).

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network (Lee and Hastie 2012). Following Faisal et al. (2010), I fit
regularized logistic regression models (Gelman et al. 2008) for each
species, using the other species on the landscape as predictors. To
avoid the identifiability problems associated with directed cyclic
graphs (Schmidt and Murphy 2012), I then symmetrized the relationships
within species pairs via averaging.

The fourth alternative method, described in Gotelli and Ulrich (2009),
involved simulating new landscapes from a null model that retains the
row and column sums of the original matrix (Strong et al. 1984). I used
the \(Z\)-scores computed by the Pairs software described in Gotelli and
Ulrich (2009) as my null model-based estimator of species interactions.

The last two estimators used the latent correlation matrix estimated by
the BayesComm package (Golding and Harris 2015) in order to evaluate the
recent claim that the correlation coefficients estimated by ``joint
species distribution models'' provide an accurate assessment of species'
pairwise interactions (Pollock et al. 2014, see also Harris 2015b). In
addition to using the posterior mean correlation (Pollock et al. 2014),
I also used the posterior mean \emph{partial} correlation, which might
be able to control for indirect effects.

@@

For the simulated landscapes based on Figure 1, I assessed whether each
method's test statistic indicated a positive or negative relationship
between the two shrubs (Appendix 1). For the null model (Pairs), I
calculated statistical significance using its \(Z\)-score. For the
Markov network, I used the Hessian matrix to generate approximate
confidence intervals and noted whether these intervals included zero.

I then evaluated the relationship between each method's estimates and
the ``true'' interaction strengths among all of the species pairs from
the larger simulated landscapes. This determined which the methods
provide a consistent way to know how strong species interactions
are---regardless of which species were present in a particular data set
or how many observations were taken. Because the different methods
mostly describe species interactions on different scales
(e.g.~correlations versus \(Z\) scores versus regression coefficients),
I used linear regression through the origin to rescale the different
estimates produced by each method so that they had a consistent
interpretation. After rescaling each method's estimates, I calculated
squared errors between the scaled interaction estimates and ``true''
interaction values across all the simulated data sets. These squared
errors determined the proportion of variance explained for different
combinations of method and landscape size (compared with a null model
that assumed all interaction strengths to be zero).

@@

As shown in Figure 1, the marginal relationship between the two shrub
species was positive---despite their competition for space at a
mechanistic level---due to indirect effects of the dominant tree
species. As a result, the correlation between these species was positive
in 94\% of replicates, and the randomization-based null model falsely
reported positive associations 100\% of the time. Worse, more than 98\%
of these false conclusions were statistically significant. The partial
correlation and Markov network estimates, on the other hand, each
correctly isolated the direct negative interaction between the shrubs
from their positive indirect interaction 94\% of the time (although the
confidence intervals overlapped zero in most replicates).

@@

Despite some variability across contexts (Figure 3A), the four methods
that controlled for indirect effects clearly performed the best: the
Markov network explained the largest portion of the variance in the
``true'' interaction coefficients (35\% overall), followed by the
generalized linear models (30\%), partial correlations from the raw
presence-absence data (28\%), and partial correlations from BayesComm,
the joint species distribution model (26\%). The benefit of choosing the
full Markov network over the other three methods was largest on the
smaller landscapes, which are also the ones that are most representative
of typical analyses in this field (Gotelli and Ulrich 2009).

The three methods that did not attempt to control for indirect
interactions all explained less than 20\% of the variance. Of these, the
sample correlation matrix based on the raw data performed the best
(19\%), followed by the null model (15\%) and BayesComm's correlation
matrix (11\%). Although these last three methods had different \(R^2\)
values, there was a close mapping among their estimates (especially
after controlling for the size of the simulated landscapes; Figure 3B).
This suggests that the effect sizes from the null model (and, to a
lesser extent, the correlation matrices from joint species distribution
models) only contain noisy versions of the same information that could
be obtained more easily and interpretably by calculating correlation
coefficients between species' presence-absence vectors.

Bootstrap resampling indicated that the above ranking of the different
methods was robust (Appendix 3). In particular, the 95\% confidence
interval of the bootstrap distribution indicated that the Markov
network's overall \(R^2\) value was between 14 and 18 percent higher
than the second-most effective method (generalized linear models) and
between 2.12 and 2.38 times higher than could be achieved by the null
model (Pairs). Bootstrap resampling of a 200-site landscape also
confirmed that the rosalia package's estimates of species' conditional
relationships were reasonably robust to sampling variation (Appendix 4).

@@

It's important to note that none of these methods can identify the exact
nature of the pairwise interactions (e.g.~which species in a
positively-associated pair is facilitating the other; Schmidt and Murphy
2012), particularly when real pairs of species can reciprocally
influence one another in multiple ways simultaneously (Bruno et al.
2003); with compositional data, there is only enough information to
provide a single number describing each species pair. To estimate
asymmetric interactions, such as commensalism or predation, ecologists
would need other kinds of data, as from time series, behavioral
observations, manipulative experiments, or natural history. These other
sources of information could also be used to augment the likelihood
function with an informative prior distribution, which could lead to
better results on some real data sets than was shown in Figure 3A.

Despite their limitations, Markov networks have enormous potential to
improve ecological understanding. In particular, they are less
vulnerable than some of the most commonly-used methods to mistakenly
identifying positive species interactions between competing species, and
can make precise statements about the conditions whre indirect
interactions will overwhelm direct ones. They also provide a simple
answer to the question of how competition should affect a species'
overall prevalence, which was a major flashpoint for the null model
debates in the 1980's (Roughgarden 1983, Strong et al. 1984). Equation 1
can be used to calculate the expected prevalence of a species in the
absence of biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee and
Hastie 2012). Competition's effect on prevalence in a Markov network can
then be calculated by subtracting this value from the observed
prevalence (cf Figure 2D). This kind of insight would have been
difficult to obtain without a generative model that makes predictions
about the consequences of species interactions; null models (which
presume \emph{a priori} that interactions do not exist) have no way to
make such predictions.

Markov networks---particularly the Ising model for binary
networks---have been studied in statistical physics for nearly a century
(Cipra 1987), and the models' properties, capabilities, and limits are
well-understood in a huge range of applications. Using the same
framework for species interactions would thus allow ecologists to tap
into an enormous set of existing discoveries and techniques for dealing
with indirect effects, stability, and alternative stable states.
Numerous other extensions are possible: for example, the states of the
interaction network can be modeled as a function of the local abiotic
environment (Lee and Hastie 2012), which would provide a rigorous and
straightforward approach to the difficult and important task of
incorporating whole networks of biotic interactions into species
distribution models (Kissling et al. 2012, Pollock et al. 2014), leading
to a better understanding of the interplay between biotic and abiotic
effects on community structure. There are even methods (Whittam and
Siegel-Causey 1981, Tjelmeland and Besag 1998) that would allow one
species to affect the sign or strength of the relationship between two
other species, tipping the balance between facilitation and exploitation
(Bruno et al. 2003).
